Mechanising Hylomorphisms for Extraction of Executable Code

### 2025-03-19 paper/paper.tex
Abstract

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion, in part due to the rigorous set of algebraic laws that
they satisfy. These laws are the basis for reasoning about program equivalences
and, therefore , they can be used for reasoning about program correctness and
optimisations.  Some of these optimisations are successfully applied by
compilers of (functional) languages.  Formalising recursion schemes in a type
theory offers additional termination guarantees, but it often requires
compromises affecting the resulting code, such as imposing performance
penalties, requiring the assumption of additional axioms, or introducing unsafe
casts into extracted code (e.g. \ocaml{Obj.magic} in OCaml).

This paper presents the first Coq formalisation of a recursion scheme, called
the \emph{hylomorphism}, along with its algebraic laws allowing for the
mechanisation of all recognised (terminating) recursive algorithms. The key
contribution of this paper is that this formalisation is fully axiom-free
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted OCaml code for the programs formalised in our
framework is efficient, resembles code that a human programmer would write, and
contains no occurrences of \ocaml{Obj.magic}.  Furthermore, we show the use of
the machine-checked proofs of the laws of hylomorphisms to do program
optimisations.
\end{abstract}

>>>>

%TODO mandatory: add short abstract of the document
\begin{abstract}
  Generic programming with recursion schemes provides a powerful abstraction for
structuring recursion and provide a rigorous reasoning principle for program
optimisations which have been successfully applied to compilers for functional
languages. Formalising recursion schemes in a type theory offers additional
termination guarantees, but it often requires compromising on performance,
requiring the assumption of additional axioms, and/or introducing unsafe casts
into extracted code.

This paper presents the first Coq formalisation of \emph{hylomorphisms} allowing for the
mechanisation of all recognised recursive algorithms. The key
contribution of this paper is that this formalisation is fully \emph{axiom-free}
allowing for the extraction of safe, idiomatic OCaml code. We exemplify the
framework by formalising a series of algorithms based on different recursive
paradigms such as divide-and conquer, dynamic programming, and mutual recursion
and demonstrate that the extracted OCaml code for the programs formalised in our
framework is efficient, humanly readable, and runnable. Furthermore, we show the use of
the machine-checked proofs of the laws of hylomorphisms to do program
optimisations.
\end{abstract}
